---
title: "Project Report: Cars Crash"
author: "Huy Ngo"
date: "Fall 2020"
output:
  pagedown::html_paged:
    toc: false
    # change to true for a self-contained document, but it'll be a litte slower for Pandoc to render
    self_contained: false
# uncomment this line to produce HTML and PDF in RStudio:
#knit: pagedown::chrome_print
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F, fig.width = 8, fig.height = 7, out.width = "100%", cache = T)
options(digits = 3)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE} 
library(dplyr)
library(readr)
library(ggplot2)
library(plotly)
library(forcats)
Crash <- read_csv("monroe-county-crash-data2003-to-2015.csv")
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(lubridate)
library(dplyr)
library(stringr)
library(plotly)
tmp <- Crash %>%
  mutate(hm = sprintf("%04d", Hour),
         hour = str_sub(hm, 1, 2) %>% parse_number(),
         minute = str_sub(hm, 3, 4) %>% parse_number()) %>%
  mutate(wday = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")[Day],
         wday_dec = Day + hour/24)
```

## Introduction

Automobiles are an indispensable part of life. Thanks to them that travel, transaction, business, and even medical problems has been improving. But they also carry risks that affect human life. People's need is higher and higher, so the number of vehicles must also increase, which entails many consequences such as traffic congestion and accidents. Every day, with hundreds of millions of car trips moving continuously across the country, the chance that one person might find him/herself get involved in an accident is not small.

Monroe County is part of the Bloomington, Indiana. It was formed in 1818 from portions of Orange County, and named after James Monroe, the fifth President of the United States (in office from 1817 - 1825). The 2010 United States Census report that the population was 137,974, the total area is 411.32 square miles (1,065.3 km2), of which 394.51 square miles (1,021.8 km2) (or 95.91%) is land and 16.81 square miles (43.5 km2) (or 4.09%) is water. The county terrain is low rolling hills, covered with vegetation and largely devoted to agricultural use or urban development. [1] 

The data use in this project is accidents reported in Monroe County from 2003 to 2015. There are more than 53.000 observations (cases reported) and many variables, like:

* Day/ Month/ Year/ Hour (that accidents occurred). 
* Collision type: How many cars involved?  
* What kind of injuries? 
* What is the primary reason that causes accidents to happen?
 

Some questions I want to find out through the data:

  - Which year has the highest rate of accident? On what month? Is that month consistent for all other years? 
  - Is there any specific day (or month, year) with more accidents than the other day (month/ year)? 
  - On that day (month/ year), what is the primary factor that causes the accident? What are the highest injuries? 
  - What can we conclude about driving habit of citizen of Monroe county?
  
## Methods

Comparing the overall with a specific year to find the "trend" of accidents.

* Find the year that has the most accidents occurred. Call it Result. 1
* Among the years, find the month that has the highest accidents. Call it Result. 2
  - Look at the year found in Result. 1, then, find the month has the most accidents. Call it Result. 3. 
  - Compare Result. 3 with Result. 2. 
* Among the years, look at the day with most accidents, identify if it weekday or weekend. Call it Result. 4
  - Look at the Result. 1 and find which type of day has more accidents. Call it Result. 5. 
  - Compare Result. 5 and Result. 4. 
* Among the years, find the hour (in a 24-hour format) with the highest accidents. Call it Result. 6
  - Look at Result. 1 and find which hour has the most accidents. Call it Result. 7
  - Compare Result. 6 and Result. 7. 
* Among the years, look for the primary factor that causes accidents. Call it Result. 8
  - Look at Result. 1 and find the primary factor for accidents. Call it Result. 9
  - Compare Result. 8 and Result. 9. 
* Look at the relationship between all the factors above. Conclude.


## Results

First, consider a frequency table below:

```{r, error = T, echo=FALSE, message=FALSE, warning=FALSE, results='markup'}
library(ggplot2)
Count<-table(tmp$Year)
knitr::kable(Count)
```

Then, consider the bars chart: 

```{r}
tmp %>%
  group_by(Year) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  ggplot(aes(x = Year, y = count)) + 
  geom_bar(stat="identity", fill="brown") + 
  geom_text(aes(label=count), vjust=1.6, color="white", size=3.5)+
  ggtitle("Figure 1. Which year has the highest accidents?") +
  theme_minimal()
  

```

Both the frequency table and the bar chart shows us that 2003 was the year with the highest number of accidents, when 2011 is has the lowest. From now we compare 2003 with the overall years. Then, we compare 2011 with the overall.


```{r, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(tmp, aes(x = Month)) + 
  geom_line(stat = "count", aes(y = ..count..))+
  geom_text(aes(Month, 6000, label=Month))+
  ggtitle("Figure 2. Accidents month by month, 2003 to 2015") +
  theme_minimal()

```

In Figure 2., we see the number of accidents peaking at two months: April and October. The peak in October is higher, indicate that month has the most accidents. From now we assume the trend "months of accidents" are April and October. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
Month<-filter(tmp, Year == "2003")
ggplot(Month, aes(x = Month)) + 
  geom_line(stat = "count", aes(y = ..count..))+
  geom_text(aes(Month, 600, label=Month))+
  ggtitle("Figure 3. Accidents month by month, 2003") +
  theme_minimal()
```

In Figure 3. we consider the year 2003 only, and we find that October is still the month with the most accidents. The 2nd place is not April but instead, February. 


```{r, echo=FALSE, message=FALSE, warning=FALSE}

tmp %>%
  group_by(wday, Year) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  ggplot(aes(x = wday, y=count, color=Year)) + 
  geom_point()+
  ggtitle("Figure 4. Accidents day by day, 2003 to 2015")

```
Figure 4. shows that Friday is the day with most accidents happened through overall years.


```{r, echo=FALSE, message=FALSE, warning=FALSE}
Month %>%
  group_by(wday, Month, Year) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  ggplot(aes(x = Month, y=count, color=wday)) + 
  geom_point()+
  ggtitle("Fig. 5. Accidents day by day, 2003")
```

In Figure 5., the red dot presents Friday. If we looked at overall months in 2003, we see the red dots appeared much time at the top of the graph, indicate that most of the accidents happened on Friday. So the data of 2003 consistent with the overall data. 

```{r, echo=FALSE, message=FALSE, warning=FALSE, figures-side, fig.show="hold", out.width="70%"}
ggplot(tmp, aes(x = hour)) + 
  geom_bar(fill="red")+
  ggtitle("Figure 6. Accidents hour by hour, 2003 to 2015") +
  theme_minimal()

ggplot(Month, aes(x = hour)) + 
  geom_bar(fill="steelblue")+
  ggtitle("Figure 7. Accidents hour by hour, 2003") +
  theme_minimal()

```

Figures 6 and 7 show us the accidents tend to happen in the daytime, starting at 7 am and peaking at 5 pm. Also, evening and night time still have a decent amount of collisions compared to the dawn time. 

Let take a look at the table of frequency accidents grouped by type of day (weekday or weekend). The table below is for the overall years 2003 to 2015. If we assume that weekdays include: Monday, Tuesday, Wednesday, Thursday, Friday, and weekend have Saturday and Sunday; thus, we have to divide the numbers in the table by five and two, accordingly. 

```{r, error = T, echo=FALSE, message=FALSE, warning=FALSE, results='markup'}
We<-table(tmp$`Weekend?`)
knitr::kable(We)

```

So, the mean accidents for weekday is: 8121.

And the mean accidents for weekend is: 6636.

We did the same thing for the table frequency 2003 below:

```{r, error = T, echo=FALSE, message=FALSE, warning=FALSE, results='markup'}
We1<-table(Month$`Weekend?`)
knitr::kable(We1)
```

The mean accidents for weekday is: 696.

The mean accidents for weekend is: 561.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

tmp %>%
  group_by(hour, `Weekend?`, Year) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  ggplot(aes(x = hour, y = count, color = `Weekend?`)) + 
  geom_line(aes(frame = Year))+
  ggtitle("Figure 8. Weekend or weekdays?")
```

We conclude that accidents tend to appear on weekdays rather than on the weekend. (Figure 8.)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(tmp, aes(x = "", fill=`Injury Type`)) + 
  geom_bar()+
  ggtitle("Figure 9. Injury types, 2003 to 2015") +
  theme_minimal()+
  coord_polar("y")
```

Next, we look at the fatal rate. There are four types of injury: Fatal, incapacitating, no injury/ unknown, and non-incapacitating.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(Month, aes(x = "", fill=`Injury Type`)) + 
  geom_bar()+
  ggtitle("Figure 10. Injury type, 2003") +
  theme_minimal()+
  coord_polar("y")
```

Luckily, most of the accidents result in no injury. The number of incapacitating and fatal cases was not high. The data looks consistent for both overall years and 2003 (Figure 9. and Figure 10.).
 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
tmp %>%
  group_by(`Collision Type`, Month) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  ggplot(aes(x = Month, y = count)) + 
  geom_col() + 
  ggtitle("Figure 11. Collision types, 2003 to 2015") +
  facet_wrap(~`Collision Type`)

```

Figure 11. and Figure 12. tell us about the type of crash: How many vehicles involved? One, two, or three? Can it be the bus? Cyclist? Motorcycle? Pedestrian? 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
tmp %>%
  filter(Year == "2003") %>%
  group_by(`Collision Type`, Month) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  ggplot(aes(x = Month, y = count)) + 
  geom_col() + 
  ggtitle("Figure 12. Collision types, 2003") +
  facet_wrap(~`Collision Type`)

```

If we look at the overall 2003 to 2015 and look at 2003 only, we see that most of the accidents are the collision between two cars.

```{r, fig.height=20, fig.width=15, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(tmp, aes(x = fct_rev(fct_infreq(`Primary Factor`)))) + 
  geom_bar()+
  ggtitle("Figure 13. Primary factors of accidents, 2003 to 2015") +
  theme_minimal()+ coord_flip()
```
For the overall years, the primary factor that for most of the accidents is "Failure to yield right of way" (Figure 13.).

```{r, fig.height=20, fig.width=15, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(Month, aes(x = fct_rev(fct_infreq(`Primary Factor`)))) + 
  geom_bar()+
  ggtitle("Figure 14. Primary factors of accidents, 2003") +
  theme_minimal()+ coord_flip()
```
For the year of 2003, "Failure to yield right of way" is still the most factor that lead to accidents (Figure 14.). It looks like the data is consistent for all of the factors. 

We can summarize some of the key-words about "trending" accidents in Monroe - county: Most accidents occurred on:

* October.
* Friday afternoon (around 5 pm). 
* Almost the cases is the crash between two cars.
* Most of it results in no injury. 
* The primary factor that cause the collisions is Failure to yield right of the way.

We have look at the year with highest accidents, now we investigate the year with lowest accidents. Recall from Figure 1., we know that years is 2011. We then look at all factors of this year: Month, day, hour, injury type, collision type, and primary factor.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
Mo<-filter(tmp, Year == "2011")
ggplot(Mo, aes(x = Month)) + 
  geom_line(stat = "count", aes(y = ..count..))+
  geom_text(aes(Month, 600, label=Month))+
  ggtitle("Figure 15. Accidents month by month, 2011") +
  theme_minimal()
```

In Figure 15., we found that October is still the month with the highest crash. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}

Mo %>%
  group_by(wday, Month, Year) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  ggplot(aes(x = Month, y=count, color=wday)) + 
  geom_point()+
  ggtitle("Fig. 16. Accidents day by day, 2011")
```

It looks like Friday is still the day with the most cases (Figure 16.). The data of 2011 looks consistent with the overall and 2003 data.


```{r, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(Mo, aes(x = hour)) + 
  geom_bar(fill="pink")+
  ggtitle("Figure 17. Accidents hour by hour, 2011") +
  theme_minimal()

```

Next, we look at the hour. Figure 17. show us the time for most of the accidents that occurred is 5 pm. The data still consistent. 


```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(Mo, aes(x = "", fill=`Injury Type`)) + 
  geom_bar()+
  ggtitle("Figure 18. Injury type, 2011") +
  theme_minimal()+
  coord_polar("y")
```
Luckily, most of the accidents result in no injury. The number of incapacitating and fatal cases was not high (Figure 18.).

```{r, echo=FALSE, message=FALSE, warning=FALSE}
tmp %>%
  filter(Year == "2011") %>%
  group_by(`Collision Type`, Month) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  ggplot(aes(x = Month, y = count)) + 
  geom_col() + 
  ggtitle("Figure 19. Collision types, 2011") +
  facet_wrap(~`Collision Type`)

```

Figure 19. let us know that most of the crashes are collisions between two cars.

```{r, fig.height=20, fig.width=15, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(Mo, aes(x = fct_rev(fct_infreq(`Primary Factor`)))) + 
  geom_bar()+
  ggtitle("Figure 20. Primary factors of accidents, 2011") +
  theme_minimal()+ coord_flip()
```

For 2011, "Failure to yield right of way" is still the most factor that lead to accidents (Figure 20.). It looks like the data is consistent for all of the factors. 


## Conclusion




## Sources
[1] https://en.wikipedia.org/wiki/Monroe_County,_Indiana
