---
title: "Project Report: Cars Crash"
author: "Huy Ngo"
date: "Fall 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 3)
```

```{r load-sas-libraries, echo = F, message = F, warning = F}

library(SASmarkdown)
sas_enginesetup(sashtml=sashtml)

sasexe <- "C:/Program Files/SASHome/SASFoundation/9.4/sas.exe"
sasopts <- "-nosplash -ls 75"

# Linux options (for grading, please leave this in!)
# sasexe <- "/usr/local/SASHome/SASFoundation/9.4/bin/sas_en"
# sasopts <- "-ls 75"
```

```{r, echo=FALSE, message=FALSE, warning=FALSE} 
library(dplyr)
library(readr)
library(ggplot2)
library(plotly)
Crash <- read_csv("monroe-county-crash-data2003-to-2015.csv")
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(lubridate)
library(dplyr)
library(stringr)
library(plotly)
tmp <- Crash %>%
  mutate(hm = sprintf("%04d", Hour),
         hour = str_sub(hm, 1, 2) %>% parse_number(),
         minute = str_sub(hm, 3, 4) %>% parse_number()) %>%
  mutate(wday = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")[Day],
         wday_dec = Day + hour/24)
```

## Introduction

* Automobiles are an indispensable part of life. Thanks to them that travel, transaction, business, and even medical problems has been improving. But they also carry risks that affect human life. People's need is higher and higher, so the number of vehicles must also increase, which entails many consequences such as traffic congestion and accidents. Every day, with hundreds of millions of car trips moving continuously across the country, the chance that one person might find him/herself get involved in an accident is not small.

* The data use in this project is accidents recorded in Monroe County, Indiana, from 2003 to 2015. There are more than 53.000 observations (accidents reported) and many variables.But I am interested in some certain variables, like:
  - Day/ Month/ Year/ Hour (that accidents occurred). 
  - How many cars involved? (Collision type). 
  - What kind of injuries? 
  - What is the primary reason that causes accidents to happen?  

```{r, engine="sashtml", engine.path=sasexe, engine.opts=sasopts, collectcode = T, error = T, echo=FALSE, message=FALSE, warning=FALSE}
ODS HTML STYLE=HTMLBLUE;
filename dat "monroe-county-crash-data2003-to-2015.csv";
PROC IMPORT DATAFILE = dat OUT = Crash
    DBMS = CSV 
    REPLACE; 
    GETNAMES = YES;
    GUESSINGROWS = 5000;
RUN;
proc print data= Crash (obs=5);
var Year Month Day Hour Collision_Type Injury_Type Primary_Factor;
title 'Table.1. Monroe County Accidents Reported from 2003 to 2015';
run;
```

* Some questions I want to find out through the data include:
  - Which year has the highest rate of accident? On what month? Is that month consistent for all other years? 
  - Is there any specific day (or month, year) with more accidents than the other day (month/ year)? 
  - On that day (month/ year), what is the primary factor that causes the accident? What are the highest injuries? 
  - What can we conclude about driving habit of citizen of Monroe county?
  
## Methods

* Comparing the overall with a specific year:
* Find the year that has the most accidents occurred. Call it Result. 1
* Among the years, find the month that has the highest accidents. Call it Result. 2
  - Look at the year found in Result. 1, then, find the month has the most accidents. Call it Result. 3. 
  - Compare Result. 3 with Result. 2. 
* Among the years, look at the day with most accidents, identify if it weekday or weekend. Call it Result. 4
  - Look at the Result. 1 and find which type of day has more accidents. Call it Result. 5. 
  - Compare Result. 5 and Result. 4. 
* Among the years, find the hour (in a 24-hour format) with the highest accidents. Call it Result. 6
  - Look at Result. 1 and find which hour has the most accidents. Call it Result. 7
  - Compare Result. 6 and Result. 7. 
* Among the years, look for the primary factor that causes accidents. Call it Result. 8
  - Look at Result. 1 and find the primary factor for accidents. Call it Result. 9
  - Compare Result. 8 and Result. 9. 
* Look at the relationship between all the factors above. Conclude.

> Note the lack of bar charts! Also, I would have expected different patterns on Saturday and Sunday, so it's a bit odd that those aren't showing up.



## Results

- The accidents recorded through year by create a frequency table. Then, create a bar graph so we can visualize it.

```{r, engine="sashtml", engine.path=sasexe, engine.opts=sasopts, collectcode = T, error = T, echo=FALSE, message=FALSE, warning=FALSE}
ODS HTML STYLE=HTMLBLUE;
title 'Table.2. Accidents frequency through years';
PROC FREQ DATA=Crash;
    TABLE Year;
output out = YE;
RUN;
Proc print data = YE;
run;
```

- From the table above, we see that the year which highest cases reported is 2003 with 4625 cases. 
- Next, we look at the overall years and see that October is the month with the most accidents.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(plotly)
p <- tmp %>%
  group_by(Month, Year) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  ggplot(aes(x = Month, y = count)) + 
  geom_line(aes(frame = Year))+
  ggtitle("Fig. 1. Overall months of accidents from 2003 to 2015") 
ggplotly(p)

```

- Look at 2003 and we find that October is still the month with most accidents (477 cases). So the data consistent.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

Month<-filter(tmp, Year == "2003")
ggplot(Month, aes(x = Month)) + 
  geom_line(stat = "count", aes(y = ..count..))+
  geom_text(aes(Month, 600, label=Month))+
  ggtitle("Fig. 2. Months of accidents in 2003") +
  theme_minimal()
```


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(plotly)

p <- tmp %>%
  group_by(wday, Year) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  ggplot(aes(x = wday, y = count)) + 
  geom_point(aes(frame = Year))+
  ggtitle("Fig. 3. Overall days of accidents from 2003 to 2015")
ggplotly(p)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(Month, aes(x = wday)) + 
  geom_point(stat = "count", aes(y = ..count..))+
  ggtitle("Fig. 4. Days of accidents in 2003") 
```



```{r, echo=FALSE, message=FALSE, warning=FALSE, figures-side, fig.show="hold", out.width="50%"}
ggplot(tmp, aes(x = `Weekend?`)) + 
  geom_bar(fill="orange")+
  ggtitle("Fig. 5. Weekday or weekend overall from 2003 to 2015") +
  theme_minimal()

ggplot(Month, aes(x = `Weekend?`)) + 
  geom_bar(fill="steelblue")+
  ggtitle("Fig. 6. Weekday or weekend in 2003") +
  theme_minimal()

```



- For the overall years, the primary factor that for most of the accidents is "Failure to yield right of way".
```{r, fig.height=20, fig.width=15, echo=FALSE, message=FALSE, warning=FALSE}
library(forcats)
ggplot(tmp, aes(x = fct_rev(fct_infreq(`Primary Factor`)))) + 
  geom_bar()+
  ggtitle("Fig. 1. Primary factors for accidents from 2003 to 2015") +
  theme_minimal()+ coord_flip()
```

- For the year of 2003, "Failure to yield right of way" is still the most factor that lead to accidents. 
```{r, fig.height=20, echo=FALSE, message=FALSE, warning=FALSE}
library(forcats)
ggplot(Month, aes(x = fct_rev(fct_infreq(`Primary Factor`)))) + 
  geom_bar()+
  ggtitle("Fig. 1. Primary factors for accidents in 2003") +
  theme_minimal()+ coord_flip()
```


```{r}
tmp %>%
  filter(Year == "2003") %>%
  group_by(`Collision Type`, Month) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  ggplot(aes(x = Month, y = count)) + 
  geom_col() + 
  ggtitle("Fig. Collision types in 2003") +
  facet_wrap(~`Collision Type`)

```



```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(plotly)

p <- tmp %>%
  group_by(hour, `Weekend?`, Year) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  ggplot(aes(x = hour, y = count, color = `Weekend?`)) + 
  geom_line(aes(frame = Year))+
  ggtitle("Fig. Accidents relate in hour and weekend overall years")
ggplotly(p)
```


```{r, engine="sashtml", engine.path=sasexe, engine.opts=sasopts, collectcode = T, error = T, echo=FALSE, message=FALSE, warning=FALSE}
ODS HTML STYLE=HTMLBLUE;
filename dat "monroe-county-crash-data2003-to-2015.csv";
PROC IMPORT DATAFILE = dat OUT = Crash
    DBMS = CSV 
    REPLACE; 
    GETNAMES = YES;
    GUESSINGROWS = 5000;
RUN;

```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(plotly)

p <- tmp %>%
  group_by(hour, wday, Year) %>%
  summarize(count = n()) %>%
  ungroup() %>%
  ggplot(aes(x = hour, y = count, color = wday)) + 
  geom_line(aes(frame = Year)) 
ggplotly(p)

ggplot(tmp, aes(x = wday_dec)) + 
  geom_line(stat = "count", aes(y = ..count..))

ggplot(tmp, aes(x = hour, color = wday)) + 
  geom_line(stat = "count", aes(y = ..count..))
```

## Conclusion
